---
title: "individual1.Rmd"
author: "Tessa Mendoza"
date: "11/3/2020"
output: html_document
---
```{r message=FALSE, warning=FALSE}
library(tidyverse)
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```

**1) Are there some days of the week where more accidents happen than the others?**

Yes - Saturday, Friday, and Sunday are the msot common days of the week for accidents.
```{r message=FALSE, warning=FALSE}
q1 <- acc %>%
  group_by(DAY_WEEK) %>%
  summarise(num = n())

ggplot(q1, aes(x = DAY_WEEK, y = num)) + geom_bar(stat = 'identity') + 
  labs(x = "Day of the Week", y = "Number of Accidents")

```

**2) Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)**

```{r}
q2 <- person %>%
  filter(INJ_SEV == 4)
head(q2)
```

**3) Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)**

```{r message=FALSE, warning=FALSE}
q3 <- q2 %>%
  filter(!is.na(MAKE)) %>%
  group_by(STATE, MAKE) %>%
  summarize(numFatalities = n()) %>%
  filter(numFatalities == max(numFatalities))
head(q3)
```

**4) Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map.**  
```{r warning=FALSE}
codes <- readxl::read_xlsx("FRPP_GLC_-_United_StatesSep292020.xlsx")
states <- map_data('state')

codes <- codes %>%
  mutate('State Code' = as.numeric(`State Code`))

q3 <- q3 %>%
  inner_join(codes %>% select("State Code", "State Name"), by = c("STATE" = "State Code")) %>%
  distinct()
q3$`State Name` <- tolower(q3$'State Name')

states <- states %>%
  left_join(q3 %>% select("MAKE", "State Name"), by = c("region"= "State Name"))

label <- states %>% 
  group_by(region) %>%
  summarise(long = mean(long), lat = mean(lat), make = min(MAKE))

label$make <- factor(label$make, levels = c(12, 20, 49, 37), 
                     labels = c("Ford", "Chevrolet", "Toyota", "Honda"))


ggplot(states %>% group_by(region), aes(x=long, y=lat)) + 
  geom_polygon(aes(group=group)) + 
  geom_text(aes(label = make), size = 2, color = 'white', data = label)

```

In this analysis of the most dangerous vehicle by state, the definition of "dangerous vehicle" was set as the number of fatalities that occurred in each vehicle model. Thus, in Iowa, the largest number of fatalities occurred in Chevrolet vehicles. From the map, it is clear that the majority of the most dangerous vehicles in each state were Ford and Chevrolet cars.

**5) Join the accident and person table (work out which variable(s) to use)**
```{r}
q5 <- person %>%
  left_join(acc)
```

The above code joins by the following columns: "STATE", "ST_CASE", "VE_FORMS", "COUNTY", "DAY", "MONTH", "HOUR", "MINUTE", "RUR_URB", "FUNC_SYS", "HARM_EV", "MAN_COLL", "SCH_BUS" to avoid duplication of columns. 

**6) Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results.**
```{r}
q6 <- q5 %>%
  filter(SEX != 8 & SEX != 9) %>%
  filter(HOUR <= 24) %>%
  group_by(DAY_WEEK, HOUR, SEX) %>%
  summarize(total = n())

q6$SEX <- factor(q6$SEX, levels = c(1, 2), 
                     labels = c("M", "F"))
q6$DAY_WEEK <- factor(q6$DAY_WEEK, labels = c("Sun", "Mon", "Tues", "Wed", "Thurs", "Fri", "Sat"))

ggplot(q6, aes(x = HOUR, y = total, fill = SEX)) + facet_grid(~DAY_WEEK) + geom_bar(stat = 'identity') +
  scale_fill_manual(values=c("#1287C9", "#AB3F54")) +
  labs(x = "Hour of the Day", y = "Number of Accidents")

```

**8) Is summer or winter more dangerous? Does this depend on states? Explore and explain.**
```{r}
winter <- acc %>%
  filter(between(MONTH, 1, 3) | MONTH == 12)

nrow(winter)
meanWinter <- nrow(winter) / 4
meanWinter
```

```{r}
summer <- acc %>%
  filter(between(MONTH, 6, 9))

nrow(summer)
meanSummer <- nrow(summer) / 4
meanSummer
```

```{r warning=FALSE}

summerByState <- summer %>%
  group_by(STATE) %>%
  summarize(summerCount = n())
summerByState

winterByState <- winter %>%
  group_by(STATE) %>%
  summarize(winterCount = n())
summerByState

byState <- summerByState %>%
  inner_join(winterByState, by = ("STATE" = "STATE"))
byState 

state <- codes %>%
  select(`State Name`, `State Code`) %>%
  inner_join(byState, by = c('State Code' = "STATE"))

state <- unique(state)
state
winterState <- state %>%
  filter (winterCount - summerCount > 0)
winterState

summerState <- state %>%
   filter (summerCount - winterCount > 0)
summerState
```

